Capabilities, MMIO and Robust Safety

### 2020-03-31 capability-meetings-notes/10-03/main.tex
MMIO writeup: started writing section on clients

\[
  \begin{array}{l}
  \X{bootSpec}(P, g, b, e, A_{\X{entry}}) \eqdef {} \\[0.8em]
    \quad
  \begin{array}{l}
    \left\{
    \begin{array}{l}
      \X{PC} \mapsto_{r} (\X{RWX}, \X{G}, 0, \X{MEM_{MAX}}, l_{\X{boot}}) \ast {} \\[0.5em]
      \bigast{r \in \X{Reg} \setminus \X{PC}}{r \mapsto_{r} \_}
      \ast \bigast{m \in \X{Mem} \setminus \MMIO}{m \mapsto_{a} \_}
      \ast \; \tracefrag{\emptyset}
    \end{array}
    \right\}
    \\[2em]
    \quad \I{Seq}\; (\I{Instr}\; \I{Executable})
    \\[0.5em]
    \left\{
    \begin{array}{l}
      \exists b_{adv}\; e_{adv}\; t.\; \\[0.5em]
      \quad
      \begin{array}{l}
        \X{PC} \mapsto_{r} (\X{RWX}, \X{G}, b_{adv}, e_{adv}, b_{adv}) \ast {} \\[0.7em]
        \bigast{a_{i} \in A_{\X{entry}}}{a_{i} \mapsto_{r} (\X{E},g,b,e,a_i)} \ast
        \bigast{r \in Reg\setminus \{\X{PC}, A_{\X{entry}}\}}{r \mapsto_{r} \X{inl}\; \_} \ast {} \\[1.4em]
        \bigast{m \in [b_{adv}, e_{adv})}{m \mapsto_{a} \X{inl}\; \_} \ast
        \bigast{m \in [b, e)}{m \mapsto_{a} \_} \ast
        \; \tracefrag{t} \ast \pure{P(t)}
      \end{array}
    \end{array}
    \right\}
  \end{array}
  \end{array}
\]

@@

\begin{mathpar}
  \mprset{vskip=0.5em}
  \inferrule
  { \forall a_{i} \in A_{\X{entry}} \ldotp\,
      \knowInv{\iota}{\exists t.\; \tracefrag{t} \ast P(t)}\!
         \vdash
         \forall W \ldotp\;(\ER^{\notMMIO} (\X{RX}, g, b, e, a_i))\spac W
         \\
   \X{bootSpec}(P,g,b,e,A_{\X{entry}}) \\
  (r_0, m_0, \emptyset) \longrightarrow^* (r, m, t)\\
  }
  {P(t)}
\end{mathpar}

@@

\subsubsection{Concrete boot and driver code}

Now we have a closer look at the concrete example we would like to verify,
including what its boot code looks like. Our goal is to prove that this concrete
example satisfies \X{bootSpec} (alternatively, \textsc{BootCodeOk}) and this
last robust safety theorem.

@@

>>>>

\[
  \begin{array}{l}
  \X{bootSpec}(P, g, b, e, A_{\X{entry}}) \eqdef {} \\[0.8em]
    \quad
  \begin{array}{l}
    \left\{
    \begin{array}{l}
      \X{PC} \mapsto_{r} (\X{RWX}, \X{G}, 0, \X{MEM_{MAX}}, l_{\X{boot}}) \ast {} \\[0.5em]
      \bigast{r \in \X{Reg} \setminus \X{PC}}{r \mapsto_{r} \_}
      \ast \bigast{m \in \X{Mem} \setminus \MMIO}{m \mapsto_{a} \_}
      \ast \; \tracefrag{\emptyset}
    \end{array}
    \right\}
    \\[2em]
    \quad \I{Seq}\; (\I{Instr}\; \I{Executable})
    \\[0.5em]
    \left\{
    \begin{array}{l}
      \exists b_{adv}\; e_{adv}\; t.\; \\[0.5em]
      \quad
      \begin{array}{l}
        \X{PC} \mapsto_{r} (\X{RWX}, \X{G}, b_{adv}, e_{adv}, b_{adv}) \ast {} \\[0.7em]
        \bigast{a_{i} \in A_{\X{entry}}}{r_{a_{i}} \mapsto_{r} (\X{E},g,b,e,a_i)} \ast
        \bigast{r \in Reg\setminus \{\X{PC}, A_{\X{entry}}\}}{r \mapsto_{r} \X{inl}\; \_} \ast {} \\[1.4em]
        \bigast{m \in [b_{adv}, e_{adv})}{m \mapsto_{a} \X{inl}\; \_} \ast
        \bigast{m \in [b, e)}{m \mapsto_{a} \_} \ast
        \; \tracefrag{t} \ast \pure{P(t)}
      \end{array}
    \end{array}
    \right\}
  \end{array}
  \end{array}
\]

@@

\begin{mathpar}
  \mprset{vskip=0.5em}
  \inferrule
  { \forall a_{i} \in A_{\X{entry}} \ldotp\,
    \knowInv{\iota}{\exists t.\; \tracefrag{t} \ast \pure{P(t)}}\!
         \vdash
         \forall W \ldotp\;(\ER^{\notMMIO} (\X{RX}, g, b, e, a_i))\spac W
         \\
   \X{bootSpec}(P,g,b,e,A_{\X{entry}}) \\
  (r_0, m_0, \emptyset) \longrightarrow^* (r, m, t)\\
  }
  {P(t)}
\end{mathpar}

@@

\subsubsection{Concrete boot and driver code}
\label{subsec:driverimpl}
Now we have a closer look at the concrete example we would like to verify,
including what its boot code looks like. Our goal is to prove that this concrete
example satisfies \X{bootSpec} (alternatively, \textsc{BootCodeOk}) and this
last robust safety theorem.

@@

\section{Potential driver clients/properties}
The concrete driver that we described in \texttt{subsec:driverimpl} does not yet enforce any useful
safety properties on the traces that it outputs.
This section describes a couple such properties that we might want to enforce,
modifications to the driver that they require and example clients that make use of our modified driver.

\subsection{Stateless properties}

\paragraph{Printing values $\le 1000$}

\paragraph{Modelling channel-specific safety properties}

\subsection{Stateful properties}

\paragraph{Performing $\le 1000$ I/O operations}

To ensure that no more than 1000 values will be read or sent, the driver needs to
keep some internal state, e.g.\ a single integer value $n$ (at location $l_n$) representing the number of times it has been called.

To accommodate this local state, the initial world will need to contain one
extra region for location $l_n$, that allows any integer values $\ge 0$. Since
this region only has a single state, we can again short-circuit the world and
enforce the constraints we want on $n$ through an invariant directly. We now
discuss the concrete constraint that we are looking for.

We have to tie this integer $n$ to the number of I/O events
present in the trace $t$, otherwise it will prove impossible to verify our
driver. Concretely, the invariant we will verify our driver under changes from
$\knowInv{\iota}{\exists t\ldotp \tracefrag{t} \ast \pure{|t| \le 1000}}$ to
$\knowInv{\iota}{\exists t\ldotp \tracefrag{t} \ast \pure{|t| \le 1000} \ast
  \exists n\ldotp l_n \mapsto n \ast n = |t| }$. Concretely, this means that
instead of having a pure predicate $P$ in our robust safety theorem, we should
also allow non-pure predicates that entail $P$, since we might have to enforce
other invariants relating to local state over the trace as well.

\begin{remark}
 I swept this under the rug in the above exposition, but if we allow boot code to run first, we have to
 make sure that the value $n$ indeed corresponds to the number of I/O events
 that have occurred when control is passed to the adversary.
\end{remark}

To make sure we can prove the driver specification, we will have to update the
robust safety theorem slightly to incorporate these changes. It will then look
as follows:

\paragraph{Never performing I/O again after some stop token has been read on $\stdin$}
